The Evolution of Phenotypically Invariant Genetic Networks

### 2017-12-01 kalman_walk_paper.tex
general edits to paper

What is known about the key quantity above, the amount of heritable variation in real regulatory networks?
%These are known at best only roughly \citep{felsenstein1988phylogenies},
%so we aim for order-of-magnitude estimates.
The coefficient $A_{ij}$ from the system \eqref{eqn:system} measures how much the rate of net production of $i$ changes
per change in concentration of $j$.
It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
than does coding sequence change affecting molecular structure \citep{schmidt2010vertebrate}.
In the context of transcription factor networks this may be affected 
not only by the binding strength of molecule $j$ to the promoter region of gene $i$
but also the effects of other transcription factors (e.g., cooperativity)
and local chromatin accessibility \citep{stefflova2013cooperativity}.
For this reason, 
the mutational target size for variation in $A_{ij}$ may be much larger than the dozens of base pairs
typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region,
and single variants may affect many entries of $\Sys$ simultaneously.
% (Recall that although these are best modeled through nonlinear terms,
% by linearizing we essentially consider first-order effects.)
% On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
% in the presence of substantial molecular noise \citep{canalization,buffering},
% suggesting that substantial variation in the micro-scale dynamics we consider here
% may be necessary to produce relevant phenotypic effects downstream.

Variation in binding site occupancy may overestimate variation in $A$,
since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin),
and variation in expression levels measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
Nonetheless,
\citet{kasowski2010variation} found differential occupancy in 7.5\%
of binding sites of a transcription factor (p65) between human individuals.
\citet{verlaan2009targeted} showed that cis-regulatory variation
accounts for around 2--6\% of expression variation in human blood-derived primary cells,
while \citep{lappalainen2013transcriptome} found that human population variation 
explained about 3\% of expression variation.
%%\plr{Get some data from at least one other species in here!}
Taken together, this suggests that variation in the entries of $\Sys$
may be on the order of at least a few percent between individuals of a population --
doubtless varying substantially between species and between genes.

>>>>

What is known about the key quantity above, the amount of heritable variation in real regulatory networks?
%These are known at best only roughly \citep{felsenstein1988phylogenies},
%so we aim for order-of-magnitude estimates.
The coefficient $A_{ij}$ from the system \eqref{eqn:system} measures how much the rate of net production of $i$ changes
per change in concentration of $j$.
It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
than does coding sequence change affecting molecular structure \citep{schmidt2010vertebrate}.
In the context of transcription factor networks this may be affected 
not only by the binding strength of molecule $j$ to the promoter region of gene $i$
but also the effects of other transcription factors (e.g., cooperativity)
and local chromatin accessibility \citep{stefflova2013cooperativity}.
For this reason, 
the mutational target size for variation in $A_{ij}$ may be much larger than the dozens of base pairs
typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region,
and single variants may affect many entries of $\allS$ simultaneously.
% (Recall that although these are best modeled through nonlinear terms,
% by linearizing we essentially consider first-order effects.)
% On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
% in the presence of substantial molecular noise \citep{canalization,buffering},
% suggesting that substantial variation in the micro-scale dynamics we consider here
% may be necessary to produce relevant phenotypic effects downstream.

Variation in binding site occupancy may overestimate variation in $A$,
since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin),
and variation in expression levels measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
Nonetheless,
\citet{kasowski2010variation} found differential occupancy in 7.5\%
of binding sites of a transcription factor (p65) between human individuals.
\citet{verlaan2009targeted} showed that cis-regulatory variation
accounts for around 2--6\% of expression variation in human blood-derived primary cells,
while \citep{lappalainen2013transcriptome} found that human population variation 
explained about 3\% of expression variation.
%%\plr{Get some data from at least one other species in here!}
Taken together, this suggests that variation in the entries of $\allS$
may be on the order of at least a few percent between individuals of a population --
doubtless varying substantially between species and between genes.
